Phytoplankton Retention Mechanisms in Estuaries: A Case Study of the Elbe Estuary

### 2023-10-04 ms.tex
submitted

Though, Estuaries are highly productive ecosystems. Their relative small area, disproportionally contributes to the global carbon cycle, along with their role as a source of nutrients and hatching grounds for marine ecosystems \citep{Cloern2014,Arevalo2023}.
% (Phytoplankton primary production in the world’s estuarine-coastal ecosystems,Fish larvae dynamics in temperate estuaries: A review on processes, patterns and factors that determine recruitment). 
While they are heavily influenced by anthropogenic stressors such as diking, dredging, and fishing, they are of tremendous importance for anthropogenic usage 
\citep{Jennerjahn2013, Brown2022, Wilson2002}.
Estuaries present challenging dynamic to their smallest residents, due to strong salinity gradient and a net transport to the ocean.
Here, we explore how phytoplankton, drifting small primary producers that form the basis of estuarine food webs, can persist within such dynamic environments.
% (Pressures, stresses, shocks and trends in estuarine ecosystems: An introduction and synthesis,
% Anthropogenic-estuarine interactions cause disproportionate greenhouse gas production: A review of the evidence base,
% Productivity, Fisheries and Aquaculture in Temperate Estuaries)

@@

Diel vertical migration is a process where organisms move up and down in the water column in response to the sun.
This movement favors retention by allowing plankton to reduce the time in the faster downstream currents at the water surface.
However, this has only been demonstrated for larger motile estuarine organisms such as zooplankton \citep{Hall2015, Kimmerer2002, Crawford1991,Hall2011}.
% (Vertical spatio-temporal patterns of phytoplankton due to migration behaviors in two shallow, microtidal estuaries: Influence on phytoplankton function and structure,
% Persistence of Tidally-oriented Vertical Migration by Zooplankton in a Temperate Estuary,
% Evidence for avoidance of flushing from an estuary by a planktonic, phototrophic ciliate, Flocculation kinetics and mechanisms of microalgae- and clay-containing suspensions in different microalgal growth phases).
Yet, a study by \citep{Anderson1985} showed that diel migrating dinoflagellates were able to out compete other non-motile phytoplankton in an embayment environment and even compensate for outwashing losses through reproduction increasing their abundance.
However, this also implies that the growing part of the population is somehow retaining their position.
If the regrowing population is also continuously drifting downstream they will not able to sustain their population in that area and ultimately die out due to unfavorable salinity conditions in marine waters \citep{Admiraal1976, vonAlvensleben2016, Jiang2020}.

Estuaries are complex and strongly dynamic systems such that it is still difficult to predict ecosystem dynamics or the effects of anthropogenic due to their complex bathymetry \citep{Michael2016, Fringer2019}. 
% (Sensitivity analysis of three-dimensional salinity simulations in North San Francisco Bay using the unstructured-grid SUNTANS model,
% The future of coastal and estuarine modeling: Findings from a workshop)
Nevertheless, previous modelling studies have investigated potential retention mechanisms of phytoplankton in estuaries.
In \citep{Simons2006,Kimmerer2014} they used a Lagrangian model to study zooplankton retention.
\citep{Simons2006} examined the dispersal and flushing time of mussel larvae in the St. Lawrence Estuary while \citep{Kimmerer2014} examined zooplankton movement in the San Francisco Estuary.
They were able to show that sinking and diel vertical migration slows the outwashing process and might be a beneficial retention strategy.
However, they did so by ignoring key processes like reproduction, mortality, and stranding or sedimentation processes.
Moreover, both studies were based on low resolution structured grid models, which we suspect to under-represent the complex bathymetry of estuarine systems \citep{Ye2018}.

Diatoms or benthic microalgae in particular have been observed to be strongly negatively buoyant, hence sinking to the riverbed and remaining there for a long time \citep{Passow1991,ThomasAnderson1998}`'
% (Species-specific sedimentation and sinking velocities of diatoms).
Studies also found sticky compounds in phytoplankton aggregates that are suspected to allow them stick to suspended particles, enabling them to sink to the riverbed or sticking to their surroundings aiding retention \citep{Kiørboe1993,VanderLee2000}.

In summary, different retention mechanisms have been observed or examined in modeling studies.
However, they did so in either in isolation in the case of the observational studies or with major simplifications in the modeling studies.
% \textit{
%     There is currently a lack of theoretical studies that allow a more comprehensive overview into the interplay of vertical migration and reproduction in combination with settling and stranding as retention mechanisms.
% }

Here, we explore possible retention mechanisms of phytoplankton using the Elbe estuary as a case study.
It is located in the north of Germany and flows into the North Sea.
Like most alluvial estuaries, it is relatively shallow, with most of it averaging only a few meters in average depth.
Similar to other European estuaries it experienced a strong anthropogenic pressure over the last centuries.
Most notably diking to restrain it to a narrow channel and dredging to improve access to the Hamburg harbor.
Unlike other major European harbors, the Hamburg harbor is located far inwards roughly 100 \unit{km} away from the coast.
To create port access the main channel experience a sudden jump in bathymetry from approximately 5 \unit{m} at border of the city to up to 20 \unit{m} in the harbor and downstream (see fig. \ref{fig:bathymetry}). 
This bathymetric jump is suspected to be the cause of a collapse of the phytoplankton , by causing an increase in oxygen depletion and high ammonium remineralization downstream of the bathymetric jump \citep{Schroeder1997,Holzwarth2018,Sanders2018}.
Ongoing dredging is being carried out to maintain the depth of the navigational channel causing high turbidity \citep{Kappenberg2001}.
While important aspects of the along-channel biochemical dynamics have been studied, little is known about their vertical and shore-to-shore dynamics \citep{Goosen1999, Dahnke2008, Sanders2018}.

@@

For this purpose, we further developed the individual-based Lagrangian model OceanTracker \citep{Vennell2021} and applied it to the Elbe estuary using the hydrodynamics calculated by a recent model SCHISM \citep{Pein2021}.
While the Lagrangian model simulated the movement of the inanimate organism, we included key phytoplankton features such as reproduction and mortality, sinking and rising, as well as diel vertical migration. % \textit{and dormancy?}
Using this model, we investigate under which conditions phytoplankton retention can be reproduced.

@@

\begin{figure}
    \includegraphics[width=12cm]{model_domain.png}
    \caption{Map of the full model domain, with Geesthacht being the upstream boarder on the right and the North-sea being the downstream border on the left. The black outline marks the edge of the model domain. Blue and green dots show an example snapshot of a fraction of the phytoplankton in the model. The location of the initial release is shown in red. Blue represents floating, green particles stranded by the receding tide. The red area is the initial release location. The background map has been provided by OpenStreetMap.}
    \label{fig:model_domain}
\end{figure}

@@

Mortality is induced either by high salinity, when they dry-out while stranded, or due to long term light limitation.
When particles are exposed to high salinity water above $20$PSU, a mortality probability of $0.5\%$ per minute is imposed.
This threshold is chosen based on a range of the salinity tolerances of estuarine phytoplankton species presented in \citep{vonAlvensleben2016}.
This is only an approximation and salinity tolerances many estuarine phytoplankton species deviate from this.
However, the main motivation for this choice is that most of the particles that die through this process have passed the isohaline for more than 12 hours, one tidal cycle, and are assumed not to return again through this isohaline.
Anything outside the 20 PSU isohaline is not considered part of the estuary for the purposes of this study.
Therefore, we are not tailoring our salinity tolerance to a specific species, but rather testing whether they can retain themselves within this isohaline.
We consider particles that are stranded outside the water by the receding tide, and lie dry for more than $7$ consecutive days to be dead and remove them.
Particles also die if they are light-limited for $28$ days \citep{Walter2017}.
They are considered light-limited at a depth of $1 \unit{m}$ based on turbidity data presented in \citep{Pein2021}.
A particle starts its life with a light budget of $28$ days, and each minute below $1 \unit{m}$ reduces this budget by $1$ minute, while the opposite applies when they are above $1 \unit{m}$.
Children of light-limited parents inherit the remaining light budget of their parents.

@@

investigate the effect of different patterns of vertical motion.
The first is monodirectional upward or downward vertical motion, representing either positively or negatively buoyant phytoplankton.
This buoyancy can be interpreted either as an active choice of buoyancy by the organism through adaptation, or as governed by the suspended matter aggregate on which the live.
For monodirectional vertical motion, we assign each particle a vertical velocity, which remains constant throughout its lifetime.
The second mode of vertical motion is diel vertical migration.
Here the particles change their direction of motion based on the current phase of the sun, creating a motion pattern where they rise during the day and sink during the night within the same velocity range.

% A large part of the Estuary is flat and regularly falls dry, making stranding and resuspension process important to predict their trajectories. 
We include a settling and resuspension model to represent tidal stranding and particles settling on the bed of the estuary.
Particles become stranded when the current grid cell becomes dry. 
They are not allowed to move from wet cells to dry cells, by the random walk dispersion applied to all particles.
A grid cell is considered dry based on the flag given in the SCHISM  hydrodynamic model output.
Once this cell is rewettet all stranded particles resuspend and are able to move again.
Particles settle on the bed once they attempt to move below the bottom model boundary and are resuspended based on a critical sheer velocity of 0.009 \unit{m s^{-1}} (see appendix for details).
The velocity profile in the bottom layer, or log layer, is calculated by
% centered equation
\begin{equation}
    U(z) = \frac{u_{*}}{\kappa} ln \frac{z}{z_{0}},
\end{equation}
where U is the friction velocity representing the drag at height z  above the seabed, $\kappa$ is the van Karman constant, $z_{0}$ is a length
scale reflecting the bottom roughness, and $u^{*}$ is the critical friction velocity.
If the friction velocity is above the critical friction velocity the particle is resuspended.
Particles that are stranded or settled on the bed are allowed to reproduce.
Particles are not only advected but also dispersed based on eddy diffusivity.
This allows us implement a dynamic dispersion that is crucial to represent tidal-pumping processes.
Dispersion was modeled using a random walk using a random number generator with a normal distribution.
Horizontally the standard distribution of the random walk was set to 0.1 \unit{m s^{-1}}.
The displacement by vertical dispersion $\partial z$ of particle $i$ is calculated by
\begin{equation}
    \partial z_{i} = K_{v}^{'}(z_{i}(n))\partial t + N(0, 2K_{v}(z_{i}))
\end{equation}
based on \citep{Yamazaki2014} where $z_{i}$ is the vertical position of the particle, $K_{v}^{'}$ is the vertical eddy diffusivity gradient, $K_{v}$ is the vertical eddy diffusivity and $N$ is the normal distribution.
The term based $K_{v}^{'}$ is needed to avoid particle accumulation on the top and bottom of the water column from the hydrodynamic model output.

@@

In the first experiment, we examine a range of different monodirectional upward or downward particle velocities from $-10$ to $+10$ \unit{mm\;s^{-1}} in $2$ \unit{mm\;s^{-1}} steps representing sinking or rising phytoplankton organisms \citep{Fennessy1996}.
\textit{
    Each vertical velocity is examined for a range of different reproduction rates, expressed as population doubling times in idealized conditions ranging from 77 to 772 days with a logarithmic scaling.
    In the following, we will use reproduction rate to refer to the prescribed population growth rate under ideal conditions and use growth rate whenever we describe population growth in nature. 
}
Hence, a total of 187 different scenarios are tested.
In the second set of model experiments, we study the influence of possible diel vertical migration patterns for the same vertical velocities and reproduction rates.

@@

\begin{figure}
    \includegraphics[width=8.3cm]{retention_success_combined.png}
    \caption[]{
        Relative population changes for the monodirectional movement (a) and diel migration (b) scenarios. Positive vertical velocities indicate an upwards drift. Positive population changes represent a retention success (green) while negative population changes represent a loss of the population (brown). 
        \textit{
            The vertical black lines indicate the boundary between successfully and unsuccessfully retained scenarios.
        }
        \texttt{
            Still playing with the vertical lines. Not happy with the current aesthetic.
        }
    }
    \label{fig:retention_success}
\end{figure}

@@

Our simulations show that the population is able to successfully retains itself under certain conditions.
Passively drifting particles are able to sustain themselves in the estuary if they have a reproduction rate that doubles their population size within approximately 4 months.
\textit{
    Note that the growth rates realized in nature may vary from this value due to e.g. nutrient or temperature limitations. The reproduction thresholds should be interpreted as an upper bound rather than an accurate estimate of the growth rate.
}
For the case of the monodirectional movement we see that a higher positive velocity (representing buoyancy) and a higher reproduction rate is more beneficial for retention success than a downward oriented velocity (sinking) and lower reproduction rates.
As expected, simulations in which the reproduction is set to zero do not show any retention success.
While it is easy to understand that high reproduction rates aid retention, we were surprised that buoyant phytoplankton particles are more successful in maintaining their growth in an estuary than sinking ones.

@@

\begin{figure}
    \includegraphics[width=8.3cm]{figures/retention_boxplot.png}
    \caption[]{Box and violin plot showing the vertical distribution of particles.  Short-living are those younger than 3 months and long-living all those older than that. Depth is measure relative to the current water surface with positive numbers being above the water surface i.e. stranded on shore.}
    \label{fig:migration-long-vs-short}
\end{figure}

@@

In this study, we investigated different strategies to explain how phytoplankton populations are able to maintain their population growth in estuaries while constantly being at risk to be transported into the open ocean.
% We find that reproduction rates in the range of XX-XX are sufficient for phytoplankton organisms to survive in the estuary under certain conditions.
% In addition, we found that buoyancy aids the retention success, whereas sinking decreases the retention success. Finally, especially shallow areas near the river banks and tidal flats represent important habitats for phytoplankton organisms and allow them to survive, whereas deeper areas lead to a faster transport of estuarine phytoplankton into the marine waters.

@@

Our results suggest that shallow areas are very important for maintaining the estuary phytoplankton population.
\textit{Plankton that consistently finds itself in areas that are dry due to the tides will regularly become stranded and therefore not move for much of the tidal cycle.}
We further see that positively buoyant plankton are more successful in retaining themselves. 
This is probably because they are more likely to be transported high up on the river bank where the water is less likely to reach them.
This effect is emphasized in flatter regions as the distance between the wash margin and constantly flooded areas is larger increasing the chance of settlement or them stranding again.

Initially, we expected sinking particles to have a higher retention success than buoyant ones. 
However, we found that faster sinking particles are less successful in retaining themselves.
Sinking velocities of less than $2 \unit{mm\;s^{-1}}$ are common for diatoms \citep{Passow1991} while larger velocities have been observed for aggregates in the Elbe estuary \citep{Fennessy1996}.
Sinking particles have a reduced downstream velocity because find themselves either settled on the riverbed not moving at all or close to the bed where the average downstream velocity is lower.
In addition, the deeper water layers of the Elbe have on average a lower downstream velocity than the upper water column or move upstream  due to temperature-induced density stratification \citep{Pein2021}.
Nevertheless, buoyant particles were more successful in their retention in our simulations.
The low chance of survival in the estuary for sinking particles might be explained by their light limitation in deeper waters.
We expected particles to die if they are exposed to dark conditions for more than 2 weeks.
Thus, sinking particles have a disadvantage to buoyant particles since they are more likely to become light limited and eventually die.
This suggests that dredging has a negative impact on sinking plankton because it increases both depth and turbidity \citep{DeJonge2014}, which increases the aphotic depth and therefore the volume of dark water relative to the amount of light water.

@@

Our model design does not resolve more complex ecosystem dynamics such as nutrient limitation and grazing by higher trophic levels. 
The Lagrangian model is performed offline, meaning it is not coupled to the Eulerian model that calculates the hydrodynamics and is performed after the fact.
Therefore, modeling the advection and dispersal of changes in concentration fields e.g. nutrients due to growth or remineralization was not easily possible.
Future modeling efforts could couple the Lagrangian model to a Eulerian model that disperses changes in concentrations fields by biotic activity throughout the model domain.
However, this would have drastically increased both developing and computational time to a point where it would have been infeasible in our time frame and due to the lack of appropriate validation data.
The key draw back of this is that growth rates could only be modeled as a constant rate in the current model description, similar to "ad libitum" experiments. 
This can lead to systematic errors in estimating population growth.
In reality, it can happen that nutrient limitation, which slows down the growth of the population, can occur, especially in the most light-saturated areas near the shore. For this reason, we may overestimate the role of shallow areas in our model.

@@

\textit{
    Our hydrodynamics data set was limited to the year 2012.
    Therefore, we were not able to study different release times with the same methodology.
    While we do not expect the general dynamics to change, future research could examine the effect of varying discharge throughout the seasons on retention.
}

@@

\begin{acknowledgements}
We thank Johannes Pein for providing and supporting the implementation of the hydrodynamic data.
Further, we thank Jana Hinners, Sina Remmers and Philipp Porada for providing helpful comments on the manuscript.
This study was funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) within the Research Training Group 2530: “Biota-mediated effects on Carbon cycling in Estuaries” (project number 407270017), contribution to Universität Hamburg and Leibniz-Institut für Gewässerökologie und Binnenfischerei (IGB) im Forschungsverbund Berlin e.V.
\end{acknowledgements}

>>>>

Estuaries are highly productive ecosystems. Their relative small area disproportionally contributes to the global carbon cycle, along with their role as a source of nutrients and hatching grounds for marine ecosystems \citep{Cloern2014,Arevalo2023}.
% (Phytoplankton primary production in the world’s estuarine-coastal ecosystems,Fish larvae dynamics in temperate estuaries: A review on processes, patterns and factors that determine recruitment). 
While they are heavily influenced by anthropogenic stressors such as diking, dredging, and fishing, they are of tremendous importance for anthropogenic usage 
\citep{Jennerjahn2013, Brown2022, Wilson2002}.
Estuaries present challenging dynamics to their smallest residents, due to strong salinity gradient and a net transport to the ocean.
Here, we explore how phytoplankton, drifting small primary producers that form the basis of estuarine food webs, can persist within such dynamic environments.
% (Pressures, stresses, shocks and trends in estuarine ecosystems: An introduction and synthesis,
% Anthropogenic-estuarine interactions cause disproportionate greenhouse gas production: A review of the evidence base,
% Productivity, Fisheries and Aquaculture in Temperate Estuaries)

@@

Diel vertical migration is a process where organisms move up and down in the water column in response to the sun.
This movement favors retention by allowing plankton to reduce the time in the faster downstream currents at the water surface.
However, this has only been demonstrated for larger motile estuarine organisms such as zooplankton \citep{Hall2015, Kimmerer2002, Crawford1991,Hall2011}.
% (Vertical spatio-temporal patterns of phytoplankton due to migration behaviors in two shallow, microtidal estuaries: Influence on phytoplankton function and structure,
% Persistence of Tidally-oriented Vertical Migration by Zooplankton in a Temperate Estuary,
% Evidence for avoidance of flushing from an estuary by a planktonic, phototrophic ciliate, Flocculation kinetics and mechanisms of microalgae- and clay-containing suspensions in different microalgal growth phases).
Yet, a study by \citet{Anderson1985} showed that diel migrating dinoflagellates were able to out compete other non-motile phytoplankton in an embayment environment and even compensate for outwashing losses through reproduction increasing their abundance.
However, this also implies that the growing part of the population is somehow retaining their position.
If the regrowing population is also continuously drifting downstream they will not able to sustain their population in that area and ultimately die out due to unfavorable salinity conditions in marine waters \citep{Admiraal1976, vonAlvensleben2016, Jiang2020}.

Estuaries are complex and strongly dynamic systems such that it is still difficult to predict ecosystem dynamics or the effects of anthropogenic impacts due to their complex bathymetry \citep{Michael2016, Fringer2019}. 
% (Sensitivity analysis of three-dimensional salinity simulations in North San Francisco Bay using the unstructured-grid SUNTANS model,
% The future of coastal and estuarine modeling: Findings from a workshop)
Nevertheless, previous modelling studies have investigated potential retention mechanisms of phytoplankton in estuaries.
In \citet{Simons2006,Kimmerer2014} they used a Lagrangian model to study zooplankton retention.
\citet{Simons2006} examined the dispersal and flushing time of mussel larvae in the St. Lawrence Estuary while() \citep{Kimmerer2014} examined zooplankton movement in the San Francisco Estuary.
They were able to show that sinking and diel vertical migration slows the outwashing process and might be a beneficial retention strategy.
However, they did so by ignoring key processes like reproduction, mortality, and stranding or sedimentation processes.
Moreover, both studies were based on low resolution structured grid models, which we suspect to under-represent the complex bathymetry of estuarine systems \citep{Ye2018}.

Diatoms or benthic microalgae in particular have been observed to be strongly negatively buoyant, hence sinking to the riverbed and remaining there for a long time \citep{Passow1991,ThomasAnderson1998}.
% (Species-specific sedimentation and sinking velocities of diatoms).
Studies also found sticky compounds in phytoplankton aggregates that are suspected to allow them stick to suspended particles, enabling them to sink to the riverbed or sticking to their surroundings aiding retention \citep{Kiørboe1993,VanderLee2000}.

In summary, different retention mechanisms have been observed or examined in modeling studies.
However, they did so either in isolation in the case of observational studies or with major simplifications in the modeling studies.
There is currently a lack of theoretical studies that allow for a more comprehensive overview into the interplay of vertical migration and reproduction in combination with settling and stranding as retention mechanisms.

Here, we explore possible retention mechanisms of phytoplankton using the Elbe estuary as a case study.
It is located in the north of Germany and flows into the North Sea.
Like most alluvial estuaries, it is relatively shallow, with most of it averaging only a few meters in average depth.
Similar to other European estuaries it experienced a strong anthropogenic pressure over the last centuries.
Most notably diking to restrain it to a narrow channel and dredging to improve access to the Hamburg harbor.
Unlike other major European harbors, the Hamburg harbor is located far inwards roughly 100 \unit{km} away from the coast.
To create port access the main channel experience a sudden jump in bathymetry from approximately 5 \unit{m} at border of the city to up to 20 \unit{m} in the harbor and downstream (see fig. \ref{fig:bathymetry}). 
This bathymetric jump is suspected to be the cause of a collapse of the phytoplankton, resulting in an increase in oxygen depletion and high ammonium remineralization downstream of the bathymetric jump \citep{Schroeder1997,Holzwarth2018,Sanders2018}.
Ongoing dredging is being carried out to maintain the depth of the navigational channel causing high turbidity \citep{Kappenberg2001}.
While important aspects of the along-channel biochemical dynamics have been studied, little is known about their vertical and shore-to-shore dynamics \citep{Goosen1999, Dahnke2008, Sanders2018}.

@@

For this purpose, we further developed the individual-based Lagrangian model OceanTracker \citep{Vennell2021} and applied it to the Elbe estuary using the hydrodynamics calculated by a recent model SCHISM \citep{Pein2021}.
While the Lagrangian model simulated the movement of the inanimate organism, we included key phytoplankton features such as reproduction and mortality, sinking and rising, as well as diel vertical migration.
Using this model, we investigate under which conditions phytoplankton retention can be reproduced.

@@

\begin{figure}
    \includegraphics[width=12cm]{model_domain_wcopyright.png}
    \caption{Map of the full model domain, with Geesthacht being the upstream boarder on the right and the North-sea being the downstream border on the left. The black outline marks the edge of the model domain. Blue and green dots show an example snapshot of a fraction of the phytoplankton in the model. The location of the initial release is shown in red. Blue represents floating, green particles stranded by the receding tide. The red area is the initial release location. The background map has been provided by OpenStreetMap.}
    \label{fig:model_domain}
\end{figure}

@@

Mortality is induced either by high salinity, when they dry-out while stranded, or due to long term light limitation.
When particles are exposed to high salinity water above $20$PSU, a mortality probability of $0.5\%$ per minute is imposed.
This threshold is chosen based on a range of the salinity tolerances of estuarine phytoplankton species presented in \citep{vonAlvensleben2016}.
This is only an approximation and salinity tolerances many estuarine phytoplankton species deviate from this.
However, the main motivation for this choice is that most of the particles that die through this process have passed the isohaline for more than 12 hours, one tidal cycle, and are assumed not to return again through this isohaline.
Anything outside the 20 PSU isohaline is not considered part of the estuary for the purposes of this study.
Therefore, we are not tailoring our salinity tolerance to a specific species, but rather testing whether they can retain themselves within this isohaline.
We consider particles that are stranded outside the water by the receding tide, and lie dry for more than $7$ consecutive days to be dead and remove them.
Particles also die if they are light-limited for $28$ days \citep{Walter2017}.
They are considered light-limited at a depth of $1 \unit{m}$ based on turbidity data presented in \citep{Pein2021}.
A particle starts its life with a light budget of $28$ days, and each minute below $1 \unit{m}$ reduces this budget by one minute, while the opposite applies when they are above $1 \unit{m}$.
Children of light-limited parents inherit the remaining light budget of their parents.

@@

We investigate the effect of different patterns of vertical motion.
The first is monodirectional upward or downward vertical motion, representing either positively or negatively buoyant phytoplankton.
This buoyancy can be interpreted either as an active choice of buoyancy by the organism through adaptation, or as governed by the suspended matter aggregate on which they live.
For monodirectional vertical motion, we assign each particle a vertical velocity, which remains constant throughout its lifetime.
The second mode of vertical motion is diel vertical migration.
Here the particles change their direction of motion based on the current phase of the sun, creating a motion pattern where they rise during the day and sink during the night within the same velocity range.

% A large part of the Estuary is flat and regularly falls dry, making stranding and resuspension process important to predict their trajectories. 
We include a settling and resuspension model to represent tidal stranding and particles settling on the bed of the estuary.
Particles become stranded when the current grid cell becomes dry. 
They are not allowed to move from wet cells to dry cells, by the random walk dispersion applied to all particles.
A grid cell is considered dry based on the flag given in the SCHISM  hydrodynamic model output.
Once this cell is rewetted all stranded particles resuspend and are able to move again.
Particles settle on the bed once they attempt to move below the bottom model boundary and are resuspended based on a critical sheer velocity of 0.009 \unit{m s^{-1}} (see appendix for details).
The velocity profile in the bottom layer, or log layer, is calculated by
\begin{equation}
    U(z) = \frac{u_{*}}{\kappa} ln \frac{z}{z_{0}},
\end{equation}
where U is the friction velocity representing the drag at height z  above the seabed, $\kappa$ is the van Karman constant, $z_{0}$ is a length
scale reflecting the bottom roughness, and $u^{*}$ is the critical friction velocity.
If the friction velocity is above the critical friction velocity the particle is resuspended.
Particles that are stranded or settled on the bed are allowed to reproduce.
Particles are not only advected but also dispersed based on eddy diffusivity.
This allows us implement a dynamic dispersion that is crucial to represent tidal-pumping processes.
Dispersion was modeled using a random walk using a random number generator with a normal distribution.
Horizontally the standard distribution of the random walk was set to 0.1 \unit{m s^{-1}}.
The displacement by vertical dispersion $\partial z$ of particle $i$ is calculated by
\begin{equation}
    \partial z_{i} = K_{v}^{'}(z_{i}(n))\partial t + N(0, 2K_{v}(z_{i}))
\end{equation}
based on \citet{Yamazaki2014} where $z_{i}$ is the vertical position of the particle, $K_{v}^{'}$ is the vertical eddy diffusivity gradient, $K_{v}$ is the vertical eddy diffusivity and $N$ is the normal distribution.
The term based $K_{v}^{'}$ is needed to avoid particle accumulation on the top and bottom of the water column from the hydrodynamic model output.

@@

In the first experiment, we examine a range of different monodirectional upward or downward particle velocities from $-10$ to $+10$ \unit{mm\;s^{-1}} in $2$ \unit{mm\;s^{-1}} steps representing sinking or rising phytoplankton organisms \citep{Fennessy1996}.
Each vertical velocity is examined for a range of different reproduction rates, expressed as population doubling times in idealized conditions ranging from 40 to 404 days with a logarithmic scaling.
In the following, we will use reproduction rate to refer to the prescribed population growth rate under ideal conditions and use growth rate whenever we describe population growth in nature. 
Hence, a total of 187 different scenarios are tested.
In the second set of model experiments, we study the influence of possible diel vertical migration patterns for the same vertical velocities and reproduction rates.

@@

\begin{figure}
    \includegraphics[width=8.3cm]{retention_success_combined.png}
    \caption[]{
        Relative population changes for the monodirectional movement (a) and diel migration (b) scenarios. Positive vertical velocities indicate an upwards drift. Positive population changes represent a retention success (green) while negative population changes represent a loss of the population (brown). 
        The vertical black lines indicate the boundary between successfully and unsuccessfully retained scenarios.
    }
    \label{fig:retention_success}
\end{figure}

@@

Our simulations show that the population is able to successfully retains itself under certain conditions.
Passively drifting particles are able to sustain themselves in the estuary if they have a reproduction rate that doubles their population size within approximately 3 months.
Note that the growth rates realized in nature may vary from this value due to e.g. nutrient or temperature limitations. The reproduction thresholds should be interpreted as an upper bound rather than an accurate estimate of the growth rate.

For the case of the monodirectional movement we see that a higher positive velocity (representing buoyancy) and higher reproduction rates are more beneficial for retention success than a downward oriented velocity (sinking) and lower reproduction rates.
As expected, simulations in which the reproduction is set to zero do not show any retention success.
While it is easy to understand that high reproduction rates aid retention, we were surprised that buoyant phytoplankton particles are more successful in maintaining their growth in an estuary than sinking ones.

@@

\begin{figure}
    \includegraphics[width=8.3cm]{retention_boxplot.png}
    \caption[]{Box and violin plot showing the vertical distribution of particles that are passively drifting.  Short-living are those younger than 3 months and long-living all those older than that. Depth is measured relative to the current water surface with positive numbers being above the water surface i.e. stranded on shore.}
    \label{fig:migration-long-vs-short}
\end{figure}

@@

In this study, we investigated different strategies to explain how phytoplankton populations are able to maintain their population growth in estuaries while constantly being at risk to be transported into the open ocean.

@@

Our results suggest that shallow areas are very important for maintaining the estuary phytoplankton population.
Plankton that consistently finds itself in areas that are dry due to the tides will regularly become stranded and therefore not move for much of the tidal cycle.
We further see that positively buoyant plankton are more successful in retaining themselves. 
This is probably because they are more likely to be transported high up on the river bank where the water is less likely to reach them.
This effect is emphasized in flatter regions as the distance between the wash margin and constantly flooded areas is larger, increasing the chance of settlement or them stranding again.

Initially, we expected sinking particles to have a higher retention success than buoyant ones. 
However, we found that faster sinking particles are less successful in retaining themselves.
Sinking velocities of less than $2 \unit{mm\;s^{-1}}$ are common for diatoms \citep{Passow1991} while larger velocities have been observed for aggregates in the Elbe estuary \citep{Fennessy1996}.
Sinking particles have a reduced downstream velocity because find themselves either settled on the riverbed not moving at all or close to the bed where the average downstream velocity is lower.
In addition, the deeper water layers of the Elbe have on average a lower downstream velocity than the upper water column or move upstream  due to temperature-induced density stratification \citep{Pein2021}.
Nevertheless, buoyant particles were more successful in their retention in our simulations.
The low chance of survival in the estuary for sinking particles might be explained by their light limitation in deeper waters.
We expected particles to die if they are exposed to dark conditions for more than two weeks.
Thus, sinking particles have a disadvantage to buoyant particles since they are more likely to become light limited and eventually die.
This suggests that dredging has a negative impact on sinking plankton because it increases both depth and turbidity \citep{DeJonge2014}, which increases the aphotic depth and therefore the volume of dark water relative to the amount of light water.

@@

Our model design does not resolve more complex ecosystem dynamics such as nutrient limitation and grazing by higher trophic levels. 
The Lagrangian model is performed offline, meaning it is not coupled to the Eulerian model that calculates the hydrodynamics and is performed after the fact.
Therefore, modeling the advection and dispersal of changes in concentration fields e.g. nutrients due to growth or remineralization was not easily possible.
Future modeling efforts could couple the Lagrangian model to a Eulerian model that disperses changes in concentrations fields by biotic activity throughout the model domain.
However, this would have drastically increased both, developing and computational time to a point where it would have been infeasible in our time frame and due to the lack of appropriate validation data.
The key draw back of this is that growth rates could only be modeled as a constant rate in the current model description, similar to "ad libitum" experiments. 
This can lead to systematic errors in estimating population growth.
In reality, it can happen that nutrient limitation, which slows down the growth of the population, can occur, especially in the most light-saturated areas near the shore. For this reason, we may overestimate the role of shallow areas in our model.

@@

Our hydrodynamics data set was limited to the year 2012.
Therefore, we were not able to study different release times with the same methodology.
While we do not expect the general dynamics to change, future research could examine the effect of varying discharge throughout the seasons on retention.

@@

\begin{acknowledgements}
We thank Johannes Pein for providing and supporting the implementation of the hydrodynamic data and Jana Hinners for her guidance through the project.
Further, we thank, Sina Remmers and Philipp Porada for providing helpful comments on the manuscript.
This study was funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) within the Research Training Group 2530: “Biota-mediated effects on Carbon cycling in Estuaries” (project number 407270017), contribution to Universität Hamburg and Leibniz-Institut für Gewässerökologie und Binnenfischerei (IGB) im Forschungsverbund Berlin e.V.
\end{acknowledgements}
